---
title: "Grassland Bird Monitoring Project - Motus Data Analyses"
author: "Sarah Endenburg"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{pdflscape}
   - \usepackage{fontspec}
   - \setmainfont{Arial}
classoption: landscape
output:
  pdf_document:
    latex_engine: xelatex
    toc: false # table of contents - I've suppressed this here and called it below so that it appears on a separate page
    toc_depth: '4' # will go 4 sections deep in the toc - there are only two in this document anyways
  html_document:
    self_contained: true
    theme: yeti
    toc: true
    toc_float: true
    smooth_scroll: true
    toc_depth: 4
    code_folding: hide
subtitle: ECCC contract no. 3000792304, deliverables 2 and 3
---
\newpage
\tableofcontents
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(motus)
library(tidyverse)
library(lubridate)
library(ggforce)
library(stringr)
library(rnaturalearth)
library(geosphere)
library(ggrepel)
library(knitr)
library(kableExtra)
library(viridis)
library(gridExtra)
library(sf)

Sys.setenv(tz = "UTC")

options(scipen = 9999)

```

# Introduction

This report presents preliminary analyses of Motus automated radio telemetry data collected for the Grassland Bird Monitoring Project (GBMP). The overarching goals of the GBMP are to monitor trends in the health and status of grassland bird populations and identify potential causes of decline to inform future conservation and management of the habitats and regions these species occupy. For migratory bird species, understanding departure dates, stopover ecology, and migratory connectivity are all crucial for informing effective habitat management where it is needed across species’ ranges

The data in this report are for six grassland songbird species: Chestnut-collared longspur (*Calcarius ornatus*), Thick-billed longspur (*Rhyncophanes mccownii*), Sprague's pipit (*Anthus spragueii*), Horned lark (*Eremophila alpestris*), Baird's Sparrow (*Centronyx bairdii*), and Grasshopper sparrow (*Ammodramus savannarum*). 

This report was generated in R version `r getRversion()`.

# Motus data cleaning summary

Table 1. Summary of GBMP Motus detections removed during the filtering process.
```{r table of detections filtered out, echo = FALSE, message = FALSE, warning = FALSE}

hitsRemovedTable <- readRDS("tableOfHitsRemoved.2025Feb11.rds")
rownames(hitsRemovedTable) <- c("Total number of hits", "Short runs", "Undeployed tags", "2021 detections", 
                                "Pre-deployment detections", "Missing tag metadata", "Duplicate rows", 
                                "Missing receiver metadata", "Noisy tower", "Outside species range", 
                                "Dropped tag", "Weird tower", "Short runs at noisy receivers", 
                                "Impossible flight speed")

hitsRemovedTable <- hitsRemovedTable |>
  select(-reasonRemoved)

kable(hitsRemovedTable, 
      format = "latex", # format as html or latex (for PDF output) - this applies to all tables
      table.attr = "style='width:80%;'",
      col.names = c("Reason removed", "Number of detections removed", "Number of detections remaining"),
      ) %>% 
  kable_styling(position='left', latex_options = "scale_down") |>
  row_spec(0, bold = TRUE)


```

# Tag deployment and detection summaries

Tags were deployed in 2022, 2023 and 2024 at the following sites:

```{r tag deployment site map, fig.cap="GBMP tag deployment locations for 2022, 2023 and 2024.", echo = FALSE, message = FALSE, warning = FALSE, results = 'hide', fig.align='left', fig.pos="H"}

tags912 <- read_csv("MotusTagsTemplate-912.csv")

tags912 <- tags912 |>
  mutate(speciesEN = case_when(
    speciesName == "Centronyx bairdii" ~ "Baird's Sparrow",
    speciesName == "Calcarius ornatus" ~ "Chestnut-collared Longspur",
    speciesName == "Eremophila alpestris" ~ "Horned Lark",
    speciesName == "Anthus spragueii" ~ "Sprague's Pipit",
    speciesName == "Ammodramus savannarum" ~ "Grasshopper Sparrow",
    speciesName == "Rhynchophanes mccownii" ~ "Thick-billed Longspur",
    TRUE ~ "NA"
  )) |>
  rename(motusTagID = tagID)

tagDepSites <- tags912 |>
  ungroup() |>
  mutate(tagDepLat = round(decimalLatitude, digits = 0),
         tagDepLon = round(decimalLongitude, digits = 0)) |>
  select(tagDepLat, tagDepLon, speciesEN, motusTagID) |>
  filter(is.na(tagDepLat) == FALSE) |>
  group_by(speciesEN, tagDepLat, tagDepLon) |>
  summarize(nTags = n_distinct(motusTagID))

world <- suppressMessages(ne_countries(scale = "medium", returnclass = "sf"))

states_Can <- suppressMessages(ne_states(country = c("canada"), returnclass = "sf"))

lakes <- suppressMessages(ne_download(scale = "large", type = 'lakes', category = 'physical',
returnclass = "sf", destdir = "mapData"))

xmax <- -90
xmin <- -120
ymax <- 60
ymin <- 45

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = lakes, fill = "lightblue3", colour = "gray44") +
  geom_sf(data = states_Can, colour = "gray44")+
  geom_sf(data = world, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  labs(x = "", y = "") +
  geom_jitter(data = tagDepSites, aes(x = tagDepLon, y = tagDepLat, fill = as.factor(speciesEN)), colour = "black", shape = 21, size = 2)+
  geom_text(data = states_Can, aes(label = postal, x = longitude, y = latitude), size = 3)+
  theme_bw()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "lightblue3")) +
  facet_wrap(~speciesEN) +
  theme(legend.position = "none") +
  scale_x_continuous(
    breaks = c(-115, -110, -105, -100, -95),  # Manually specify the breaks (in degrees)
    labels = c("115°W", "110°W", "105°W", "100°W", "95°W")  # Custom labels for the breaks
  ) +
  scale_y_continuous(
    breaks = c(46, 50, 54, 58)
  ) +
  # Rotate x-axis labels
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e"))

```
 
\subsection{Tag deployment and detection summaries, by species and year}

Let's get an idea of how many tags were deployed, by year and species, and how many of those were subsequently detected by the Motus network. These data have been filtered to remove detections that are likely to be false.



Table 2. Number of GBMP Motus tags deployed and detected, by species and year. 'Detections' refer to individual bursts from a tag, and the number of days detected was calculated as the number of days between deployment and the last detection for that tag.
```{r Read in Motus detections data, class.source = "fold-show", echo = FALSE, message = FALSE, fig.align='left', fig.pos="H"}
df.alltags.912.clean <- readRDS("df.alltags.912.clean.2025Feb11.rds")

df.alltags.912.clean <- df.alltags.912.clean |>
  group_by(motusTagID) |>
  mutate(numDaysDetected = as.numeric(difftime(max(ts), min(ts), units = "days")),
         nHits = n(),
         nRecv = n_distinct(recvDeployName))

detections_by_Species_Year <- df.alltags.912.clean |>
  select(speciesEN, tagDepYear, motusTagID, nHits, numDaysDetected, nRecv) |>
  distinct() |>
  ungroup() |>
  group_by(tagDepYear, speciesEN) |>
  summarise(nTagsDetected = n_distinct(motusTagID),
            nHitsAllTags = sum(nHits),
            avgHitsPerTag = nHitsAllTags/nTagsDetected,
            avgHitsPerTag_SD = sd(nHits, na.rm = TRUE),
            avgDaysDetected = mean(numDaysDetected),
            avgDaysDetected_SD = sd(numDaysDetected, na.rm = TRUE),
            meanNRecv = mean(nRecv),
            nRecv_sd = sd(nRecv),
            .groups = 'drop') |>
  arrange(speciesEN, tagDepYear) |>
  mutate(
    avgHitsPerTag = paste0(round(avgHitsPerTag, 0), " ± ", round(avgHitsPerTag_SD, 0)),  # Combine mean and SD
    avgDaysDetected = paste0(round(avgDaysDetected, 0), " ± ", round(avgDaysDetected_SD, 0)),  # Combine mean and SD
    meanNRecv = paste0(round(meanNRecv, 0), " ± ", round(nRecv_sd, 0))
  ) |>
  select(-avgHitsPerTag_SD, -avgDaysDetected_SD, -nRecv_sd)

# adding in number of tags deployed per year and species, to compare with number detected

tagDeps <- tags912 |>
  filter(is.na(utcYearStart) == FALSE) |>
  rename(tagDepYear = utcYearStart) |>
  group_by(tagDepYear, speciesEN) |>
  summarise(nTagsDeployed = n())

depTags912 <- tags912 |>
  filter(is.na(utcYearStart) == FALSE)

detections_by_Species_Year <- detections_by_Species_Year |>
  left_join(tagDeps, by = c("tagDepYear", "speciesEN")) |>
  mutate(percentTagsDetected = (nTagsDetected/nTagsDeployed)*100) |>
  dplyr::select(tagDepYear, speciesEN, nTagsDeployed, nTagsDetected, percentTagsDetected, nHitsAllTags, avgHitsPerTag, avgDaysDetected, meanNRecv) |>
 mutate(across(where(is.numeric), ~ round(., 0))) |>
  arrange(speciesEN, tagDepYear) |>
  mutate(speciesEN = as.factor(speciesEN))

species_vector <- detections_by_Species_Year$speciesEN

kable(detections_by_Species_Year, 
      format = "latex",
      table.attr = "style='width:100%;'",
      col.names = c("Year", "Species", "Number of tags deployed", "Number of tags detected", "Percent tags detected", "Total number of detections", "Average number of detections per tag", "Average number of days detected", "Number of receivers with detections")
) %>% 
  kable_styling(position='left', latex_options = "scale_down") |>
  row_spec(0, bold = TRUE)
```

Note that there are other confounding factors that likely contributed to the number of detections, including tag type.

\subsection{Detections by tag type}

Five types of tags were deployed for the GBMP:


Table 3. GBMP tag types by species.
```{r table of tag types, echo = FALSE, message = FALSE, fig.align='left', fig.pos="H"}

tags <- read_csv("tags.csv")

tagTypes <- tags |>
  select(tagID, model, motusScientificName) |>
  mutate(speciesEN = case_when(
    motusScientificName == "Calcarius ornatus" ~ "Chestnut-collared Longspur",
    motusScientificName == "Anthus spragueii" ~ "Sprague's Pipit",
    motusScientificName == "Rhynchophanes mccownii" ~ "Thick-billed Longspur",
    motusScientificName == "Eremophila alpestris" ~ "Horned Lark", 
    motusScientificName == "Centronyx bairdii" ~ "Baird's Sparrow",
    motusScientificName == "Ammodramus savannarum" ~ "Grasshopper Sparrow",
    TRUE ~ "NA")) |>
  select(-motusScientificName) |>
  rename(motusTagID = tagID,
         tagModel = model) |>
  filter(motusTagID %in% depTags912$motusTagID)|>
  group_by(tagModel) |>
  summarise(nTags = n_distinct(motusTagID),
            HOLA = sum(speciesEN == "Horned Lark"),
            CCLO = sum(speciesEN == "Chestnut-collared Longspur"),
            SSPI = sum(speciesEN == "Sprague's Pipit"),
            BAIS = sum(speciesEN == "Baird's Sparrow"),
            GRSP = sum(speciesEN == "Grasshopper Sparrow"),
            TBLO = sum(speciesEN == "Thick-billed Longspur")) |>
  mutate(tagManufacturer = case_when(
    tagModel %in% c("LifeTag", "HybridTag") ~ "CTT",
    TRUE ~ "Lotek"
  )) |>
  select(tagManufacturer, tagModel, nTags, BAIS, CCLO, GRSP, HOLA, SSPI, TBLO)

df_total <- tagTypes %>%
  summarise(
    tagModel = "Total",  # Label for the total row
    nTags = sum(nTags),
    BAIS = sum(BAIS),
    CCLO = sum(CCLO),
    GRSP = sum(GRSP),
    HOLA = sum(HOLA),
    SSPI = sum(SSPI),
    TBLO = sum(TBLO)
  )

tagTypes_with_total <- bind_rows(tagTypes, df_total) |>
  mutate(across(1, ~ if_else(is.na(.), "", .)))  

kable(tagTypes_with_total, 
      format = "latex",
      table.attr = "style='width:80%;'",
      col.names = c("Manufacturer", "Tag model", "Total number of tags", "Baird's Sparrow", "Chestnut-collared Longspur", "Grasshopper Sparrow", "Horned Lark", "Sprague's Pipit", "Thick-billed Longspur")
) %>% 
  kable_styling(position='left', latex_options = "scale_down") |>
  row_spec(nrow(tagTypes_with_total), bold = TRUE) %>%
  row_spec(0, bold = TRUE) |>
  # Optionally, add other styling for borders or layout adjustments
  row_spec(nrow(tagTypes_with_total), extra_css = "border-bottom: 2px solid #e0e0e0;") |>
  row_spec(nrow(tagTypes_with_total) -1, extra_css = "border-bottom: 2px solid #e0e0e0;")

```


CTT LifeTags are solar-powered, whereas the HybridTags use a combination of battery and solar power. Lotek tags are all battery-operated but the different models have different battery sizes (and therefore, expected lifetimes).

Did the number of detections differ by tag type?


Table 4. Detections by tag model. There were no detections for any of the four Lotek NTQB2-3-2-M tags.
```{r table of detections by tag type, echo = FALSE, message = FALSE, fig.align='left', fig.pos="H"}

detections_by_tagModel <- df.alltags.912.clean |>
  ungroup() |>
  select(tagModel, motusTagID, nHits, numDaysDetected, nRecv) |>
  distinct() |>
  group_by(tagModel) |>
  summarise(nTagsDetected = n_distinct(motusTagID),
            nHitsAllTags = n(),
            avgHitsPerTag = mean(nHits),
            avgHitsPerTag_SD = sd(nHits, na.rm = TRUE),
            avgDaysDetected = mean(numDaysDetected),
            avgDaysDetected_SD = sd(numDaysDetected, na.rm = TRUE),
            meanNRecv = mean(nRecv),
            nRecv_sd = sd(nRecv),
            .groups = 'drop') |>
  mutate(
    avgHitsPerTag = paste0(round(avgHitsPerTag, 0), " ± ", round(avgHitsPerTag_SD, 0)),  # Combine mean and SD
    avgDaysDetected = paste0(round(avgDaysDetected, 0), " ± ", round(avgDaysDetected_SD, 0)),  # Combine mean and SD
    meanNRecv = paste0(round(meanNRecv, 0), " ± ", round(nRecv_sd, 0))
  ) |>
  select(-avgHitsPerTag_SD, -avgDaysDetected_SD, -nRecv_sd)

tagModels <- tags |>
  select(tagID, model) |>
  rename(motusTagID = tagID,
         tagModel = model) |>
  filter(motusTagID %in% depTags912$motusTagID)

depTagModels <- tagModels |>
  group_by(tagModel) |>
  summarise(nTagsDeployed = n())

detections_by_tagModel <- detections_by_tagModel |>
  left_join(depTagModels, by = "tagModel") |>
  select(tagModel, nTagsDeployed, nTagsDetected, nHitsAllTags, avgHitsPerTag, avgDaysDetected, meanNRecv) |>
  mutate(percentTagsDetected = (nTagsDetected/nTagsDeployed)*100) |>
  select(tagModel, nTagsDeployed, nTagsDetected, percentTagsDetected, nHitsAllTags, avgHitsPerTag, avgDaysDetected, meanNRecv) |>
  mutate(across(where(is.numeric), ~ round(., 1)))

kable(detections_by_tagModel,
      format = "latex",
      table.attr = "style='width:100%;'",
      col.names = c("Tag model", "Number of tags deployed", "Number of tags detected", "Percent tags detected", "Total number of detections", "Average number of detections per tag", "Average number of days detected per tag", "Number of receivers with detections")
) %>% 
  kable_styling(position='left', latex_options = "scale_down") |>
  row_spec(0, bold = TRUE)

```

\subsection{Detections by harness type}

Some tags were deployed with tubes on the harness while others had no tubes. The tables below show detections for tube vs. no-tube tagged birds, first for all tags and then for Lotek 2-5-1 tags only (since all of the Lotek 2-2 tags were no-tube and also had a shorter battery life, which would impact detections). Note though that all of the 2-5-1 tags deployed at Suffield had no tubes, so that is also a confounding factor because there was no local receiver at that site, and tags were less likely to be detected as a result.

Table 5. Detections by harness type (tube or no-tube) for all deployed GBMP tags.
```{r tube vs. no tube table, echo = FALSE, message = FALSE, fig.align='left', fig.pos="H"}

bandingData <- read_csv("GBMP_and_Suffield_banding_data_Motusonly.csv")

# selecting the columns I need
harnessData <- bandingData |>
  select(Year, Month, Day, Species_Code, Tag_Type, Tag_ID, tube_notube)

# now need to match to motusTagID based on deployment date, species and tagID (there is no motusTagID in the banding data)
harnessData <- harnessData |>
  mutate(tagDeployDate = as.Date(paste(Year, Month, Day, sep = "-")),
         tube = case_when(tube_notube == "no_tube" ~ "n",
                          tube_notube == "tube" ~ "y",
                          TRUE ~ "NA")) |>
  select(-Year, - Month, -Day, -Tag_Type, -tube_notube)

# need to round mfgID because some of them have decimals but the IDs in the harness data do not
df.alltags.912.clean <- df.alltags.912.clean %>%
  ungroup() |>
  mutate(mfgID_round = ifelse(grepl("^[-+]?[0-9]*\\.?[0-9]+$", mfgID), 
                             round(as.numeric(mfgID), 0), 
                             mfgID))
# get a warning about NAs introduced by coercion but it appears to have worked

harnessData <- harnessData |>
  rename(mfgID_round = Tag_ID,
         bandingCode = Species_Code)

df.alltags.912.clean <- df.alltags.912.clean |>
  mutate(bandingCode = case_when(
    speciesEN == "Sprague's Pipit" ~ "SPPI",
    speciesEN == "Chestnut-collared Longspur" ~ "CCLO",
    speciesEN == "Grasshopper Sparrow" ~ "GRSP",
    speciesEN == "Thick-billed Longspur" ~ "TBLO",
    speciesEN == "Horned Lark" ~ "HOLA",
    speciesEN == "Baird's Sparrow" ~ "BAIS",
    TRUE ~ NA
  ),
  tagDeployDate = as.Date(tagDeployDate))

# adding to the detections data
df.alltags.912.clean <- df.alltags.912.clean |>
  left_join(harnessData, by = c("mfgID_round", "bandingCode", "tagDeployDate"))

df.alltags.912.clean <- df.alltags.912.clean |>
  group_by(motusTagID) |>
  mutate(numDaysDetected = as.numeric(difftime(max(ts), min(ts), units = "days")),
         nHits = n(),
         nRecv = n_distinct(recvDeployName))

# summary of detected tags, by harness type
detections_by_harness_type <- df.alltags.912.clean |>
  ungroup() |>
  filter(!tube == "NA" & !is.na(tube)) |>
  group_by(tube, speciesEN) |>
  summarise(nTagsDetected = n_distinct(motusTagID),
            nHitsAllTags = n(),
            avgHitsPerTag = mean(nHits),
            avgHitsPerTag_SD = sd(nHits, na.rm = TRUE),
            avgDaysDetected = mean(numDaysDetected),
            avgDaysDetected_SD = sd(numDaysDetected, na.rm = TRUE),
            meanNRecv = mean(nRecv),
            nRecv_sd = sd(nRecv),
            nTagsInNextYear = n_distinct(motusTagID[year == as.numeric(tagDepYear + 1)]),
            .groups = 'drop') |>
  mutate(
    avgHitsPerTag = paste0(round(avgHitsPerTag, 0), " ± ", round(avgHitsPerTag_SD, 0)),  # Combine mean and SD
    avgDaysDetected = paste0(round(avgDaysDetected, 0), " ± ", round(avgDaysDetected_SD, 0)),  # Combine mean and SD
    meanNRecv = paste0(round(meanNRecv, 0), " ± ", round(nRecv_sd, 0))
  ) |>
  select(-avgHitsPerTag_SD, -avgDaysDetected_SD, -nRecv_sd) |>
  arrange(speciesEN, tube)

# just for Lotek 2-5-1 nanotags
detections_by_harness_type_251 <- df.alltags.912.clean |>
  ungroup() |>
  filter(!tube == "NA" & !is.na(tube) & tagModel == "NTQB2-5-1-M") |>
  group_by(tube, speciesEN) |>
  summarise(nTagsDetected = n_distinct(motusTagID),
            nHitsAllTags = n(),
            avgHitsPerTag = mean(nHits),
            avgHitsPerTag_SD = sd(nHits, na.rm = TRUE),
            avgDaysDetected = mean(numDaysDetected),
            avgDaysDetected_SD = sd(numDaysDetected, na.rm = TRUE),
            meanNRecv = mean(nRecv),
            nRecv_sd = sd(nRecv),
            nTagsInNextYear = n_distinct(motusTagID[year == as.numeric(tagDepYear + 1)]),
            .groups = 'drop') |>
  mutate(
    avgHitsPerTag = paste0(round(avgHitsPerTag, 0), " ± ", round(avgHitsPerTag_SD, 0)),  # Combine mean and SD
    avgDaysDetected = paste0(round(avgDaysDetected, 0), " ± ", round(avgDaysDetected_SD, 0)),  # Combine mean and SD
    meanNRecv = paste0(round(meanNRecv, 0), " ± ", round(nRecv_sd, 0))
  ) |>
  select(-avgHitsPerTag_SD, -avgDaysDetected_SD, -nRecv_sd) 

# now, how many of each harness type were deployed
tube_deps <- harnessData |>
  filter(!tube == "NA" & !is.na(tube)) |>
  mutate(speciesEN = case_when(bandingCode == "SPPI" ~ "Sprague's Pipit",
                               bandingCode == "CCLO" ~ "Chestnut-collared Longspur",
                               bandingCode == "BAIS" ~ "Baird's Sparrow",
                               bandingCode == "HOLA" ~ "Horned Lark", 
                               bandingCode == "GRSP" ~ "Grasshopper Sparrow",
                               bandingCode == "TBLO" ~ "Thick-billed Longspur",
                               TRUE ~ "NA")) |>
  group_by(tube, speciesEN) |>
  summarise(nTagsDeployed = n())
  
# combining deployments with detections for each harness type
detections_by_harness_type <- tube_deps |>
  left_join(detections_by_harness_type, by = c("tube", "speciesEN")) |>
  mutate(percentTagsDetected = round((nTagsDetected/nTagsDeployed)*100, digits = 0),
         percentTagsDetectedNextYear = round((nTagsInNextYear/nTagsDeployed)*100, digits = 0)) |>
  select(tube, speciesEN, nTagsDeployed, nTagsDetected, percentTagsDetected, nHitsAllTags, avgHitsPerTag, avgDaysDetected, meanNRecv, nTagsInNextYear, percentTagsDetectedNextYear) |>
  arrange(speciesEN, tube)

# banding data doesn't have tag model in there, so need to use the tag file that I loaded previously that has that info to look just at # 2-5-1 tags deployed with and without harnesses...
deployed_tags <- tags |>
  select(mfgID, dtStart, motusScientificName, model) |>
  filter(is.na(dtStart) == FALSE) |>
  mutate(mfgID_round = ifelse(grepl("^[-+]?[0-9]*\\.?[0-9]+$", mfgID), 
                              round(as.numeric(mfgID), 0), 
                              mfgID),
         tagDeployDate = as.Date(dtStart),
         bandingCode = case_when(
           motusScientificName == "Anthus spragueii" ~ "SPPI",
           motusScientificName == "Calcarius ornatus" ~ "CCLO",
           motusScientificName == "Rhynchophanes mccownii" ~ "TBLO",
           motusScientificName == "Eremophila alpestris" ~ "HOLA",
           motusScientificName == "Centronyx bairdii" ~ "BAIS",
           motusScientificName == "Ammodramus savannarum" ~ "GRSP",
           TRUE ~ NA),
         speciesEN = case_when(
           motusScientificName == "Anthus spragueii" ~ "Sprague's Pipit",
           motusScientificName == "Calcarius ornatus" ~ "Chestnut-collared Longspur",
           motusScientificName == "Rhynchophanes mccownii" ~ "Thick-billed Longspur",
           motusScientificName == "Eremophila alpestris" ~ "Horned Lark",
           motusScientificName == "Centronyx bairdii" ~ "Baird's Sparrow",
           motusScientificName == "Ammodramus savannarum" ~ "Grasshopper Sparrow",
           TRUE ~ NA
         )) |>
  select(-mfgID, -dtStart, -motusScientificName)

tube_deps_251 <- harnessData |>
  left_join(deployed_tags, by = c("bandingCode", "tagDeployDate", "mfgID_round")) |>
  filter(model == "NTQB2-5-1-M") |>
  group_by(tube, speciesEN) |>
  summarise(nTagsDeployed = n())

detections_by_harness_type_251 <- tube_deps_251 |>
  left_join(detections_by_harness_type_251, by = c("tube", "speciesEN")) |>
  mutate(percentTagsDetected = round((nTagsDetected/nTagsDeployed)*100, digits = 0),
         percentTagsDetectedNextYear = round((nTagsInNextYear/nTagsDeployed)*100, digits = 0)) |>
  select(tube, speciesEN, nTagsDeployed, nTagsDetected, percentTagsDetected, nHitsAllTags, avgHitsPerTag, avgDaysDetected, meanNRecv, nTagsInNextYear, percentTagsDetectedNextYear) |>
  arrange(speciesEN, tube)

kable(detections_by_harness_type,
      format = "latex",
      table.attr = "style='width:100%;'",
      col.names = c("Tube", "Species", "Number of tags deployed", "Number of tags detected", "Percent tags detected", "Total number of detections (all tags)", "Average number of detections per tag", "Average number of days detected per tag", "Average number of receivers with detections", "Number of tags with detections the following year", "Percent tags with detections the following year")
) %>% 
  kable_styling(position='left', latex_options = "scale_down") |>
  row_spec(0, bold = TRUE)

```

Table 6. Detections by harness type (tube or no-tube) for Lotek 2-5-1 tags only.
```{r tube vs. no tube table 2-5-1, echo = FALSE, message = FALSE, fig.align='left', fig.pos="H"}

kable(detections_by_harness_type_251,
      format = "latex",
      table.attr = "style='width:100%;'",
      col.names = c("Tube", "Species", "Number of tags deployed", "Number of tags detected", "Percent tags detected", "Total number of detections (all tags)", "Average number of detections per tag", "Average number of days detected per tag", "Average number of receivers with detections", "Number of tags with detections the following year", "Percent tags with detections the following year")
) %>% 
  kable_styling(position='left', latex_options = "scale_down") |>
  row_spec(0, bold = TRUE)

```

## Detections vs. distance between tag deployment locations and closest receivers
The distance between tag deployment locations and receiver locations may also have impacted the number of detections because receivers have limited detection distances. The typical maximum detection distance for a Yagi antenna is about 15km, although the actual detection distance depends on factors including local topography and vegetation. For these analyses, I used the distance to the closest receiver that was active by July 31 of the year each tag was deployed rather than only looking at receivers that were active on specific tag deployment dates. Some receivers were set up near tagging locations within a few weeks of tag deployment, so using the exact deployment dates yielded some misleading results.

```{r prep code for figures for # detections vs. deployment distance from tower, echo = FALSE, message = FALSE}

depTags912 <- depTags912 |>
  select(motusTagID, mfgID, period, utcYearStart, utcMonthStart, utcDayStart, utcHourStart, utcMinuteStart, decimalLatitude, decimalLongitude) |>
  mutate(tagDeployDate = make_datetime(
    utcYearStart,
    utcMonthStart,
    utcDayStart
  )) |>
  rename(tagBI = period,
         tagDepLat = decimalLatitude,
         tagDepLon = decimalLongitude) |>
  filter(is.na(tagDepLat) == FALSE) # there are two tags with no deployment lat/long, waiting for those from Nancy

allRecv <- read_csv("receiver-deployments.csv")

allRecv$dtStart <- as.POSIXct(allRecv$dtStart)
allRecv$dtEnd <- as.POSIXct(allRecv$dtEnd)

activeRecv <- allRecv |>
  filter((dtEnd > "2022-06-01 00:00:00 UTC" | is.na(dtEnd)) & dtStart < "2024-12-01 00:00:00 UTC") |>
  rename(recvDeployName = deploymentName,
         recvDeployLat = latitude,
         recvDeployLon = longitude)


allRecv <- allRecv |>
  rename(recvDeployName = deploymentName,
         recvDeployLat = latitude,
         recvDeployLon = longitude)

activeRecv <- allRecv |>
  filter((dtEnd > "2022-06-01 00:00:00 UTC" | is.na(dtEnd)) & dtStart < "2024-12-01 00:00:00 UTC") 
# didn't have the option of an NA end date in there before, need to fix that in 4a

get_active_receivers <- function(tagDeployDate) {
  
  tagDepYear <- format(tagDeployDate, "%Y")
  tagDepMonth <- format(tagDeployDate, "%m")
  
  end_of_july <- as.Date(paste(tagDepYear, "07", "31", sep = "-"))
  
  active_receivers <- activeRecv %>% # have to have recvDeployName in there, it's not in allRecv
    filter(dtStart <= end_of_july & (dtEnd >= end_of_july | is.na(dtEnd))) %>% # recv that are still active have an NA for their end date!
    select(recvDeployName, recvDeployLat, recvDeployLon, dtStart, dtEnd)
  return(active_receivers)
}

# Step 2: Calculate the distance between tag deployments and active receivers
# A function to calculate the closest receiver using the Haversine formula

calculate_closest_receiver <- function(tag_row) {
  
  # Get the active receivers at the time of tag deployment
  active_receivers <- get_active_receivers(tag_row$tagDeployDate)
  
  # Calculate distances to all active receivers
  distances <- distVincentyEllipsoid(cbind(active_receivers$recvDeployLon, active_receivers$recvDeployLat),
                                     c(tag_row$tagDepLon, tag_row$tagDepLat))
  
  # Find the closest receiver
  closest_idx <- which.min(distances)
  
  # Return the closest receiver information
  return(data.frame(tag_id = tag_row$motusTagID, 
                    closest_receiver = active_receivers$recvDeployName[closest_idx],
                    distance_km = distances[closest_idx] / 1000))  # convert meters to km
}

closest_receivers <- do.call(rbind, lapply(1:nrow(depTags912), function(i) {
  tag_row <- depTags912[i, ] # get the current tag's row
  calculate_closest_receiver(tag_row) # pass the tag row to the function
}))

# Combine the results

closest_receivers <- closest_receivers |>
  rename(motusTagID = tag_id,
         distToClosestRecv = distance_km)

detections_deploy_dist_from_recv <- df.alltags.912.clean |>
  ungroup() |>
  group_by(speciesEN, tagModel, motusTagID) |>
  summarise(nHits = n(),
            numDaysDetected = mean(numDaysDetected),
            nRecv = n_distinct(recvDeployName),
            .groups = 'drop')

detections_deploy_dist_from_recv <- detections_deploy_dist_from_recv |>
  left_join(closest_receivers, by = "motusTagID")

tagBrands <- tags |>
  select(tagID, manufacturer) |>
  rename(motusTagID = tagID,
         tagBrand = manufacturer)

detections_deploy_dist_from_recv <- detections_deploy_dist_from_recv |>
  left_join(tagBrands, by = "motusTagID")

```

```{r figure for recv distance to tag dep site vs nHits, fig.cap="Number of detections per tag as a function of the distance (km) from tag deployment site to the closest active receiver.", fig.align='left', fig.width=10, fig.height=6, fig.pos="H"}
ggplot(detections_deploy_dist_from_recv) +
  geom_point(aes(x = distToClosestRecv, y = nHits, fill = as.factor(speciesEN), shape = as.factor(tagBrand)), size = 4, alpha = 0.5, colour = "black", position = position_jitter(width = 0.2, height = 0.2)) +
  theme_classic() +
  labs(x = "Distance from tag deployment site to closest active receiver (km)", y = "Number of detections", shape = "Tag Brand") +
  facet_wrap(~speciesEN, scales = "free_y") + 
  scale_shape_manual(values = c(21, 24)) +
  theme(axis.title.x = element_text(size = 11), 
    axis.title.y = element_text(size = 11),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 11)) +
  guides(fill = "none") +
  scale_fill_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e"))

```

```{r figure for recv distance to tag dep site vs nRecv with detections, fig.cap="Number of receivers with detections per tag as a function of the distance (km) from tag deployment site to the closest active receiver.", fig.align='left', fig.width=10, fig.height=6, fig.pos="H"}
ggplot(detections_deploy_dist_from_recv) +
  geom_point(aes(x = distToClosestRecv, y = nRecv, fill = speciesEN, shape = as.factor(tagBrand)), size = 4, alpha = 0.5, colour = "black", position = position_jitter(width = 0.2, height = 0.2)) +
  theme_classic() +
  labs(x = "Distance from tag deployment site to closest active receiver (km)", y = "Number of receivers with detections", shape = "Tag Brand") +
  facet_wrap(~speciesEN, scales = "free_y") + 
  scale_shape_manual(values = c(21, 24)) +
  theme(axis.title.x = element_text(size = 11), 
    axis.title.y = element_text(size = 11),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 11)) +
  guides(fill = "none") +
  scale_fill_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e"))

```

# Departure dates

## Fall departure dates

It is tricky to determine migratory departure dates using Motus data because there can be long gaps in between detections, and tag loss/mortality can look just like a migratory departure if the bird is not detected again post-departure. Furthermore, the six species tagged for the GBMP seemed to around on the breeding grounds before actually leaving on migration, sometimes up to ~150km. 

I calculated fall departure dates as the last detection for each bird within 100km of the tagging location, and then classified those dates based on how confident I am that each one represents an actual migratory departure. I based this on the expected timing of departure (from Birds of the World and literature) as well as the amount of time between the apparent departure and the next detection (or if the bird was detected at all post-departure):

**Likely migratory departure**: the departure date was in August, September or October and the individual was detected again moving south within 2 weeks post-departure

**Migratory departure or regional movement**: the departure date was in August, September or October but there were more than 2 weeks between the departure and the next detection

**Possible migratory departure**: the departure date was in August, September or October but the individual was not detected again post-departure, so this may also represent tag loss/malfunction

**Regional movement**: the departure date was in June or July, but the bird was detected later on so we know the tag was not lost/malfunctioning

**Probable tag loss**: the departure date was in June or July and the bird was NOT detected again

```{r fall departure dates, echo = FALSE, warning = FALSE, message = FALSE}

df.alltags.912.clean <- df.alltags.912.clean |> 
  ungroup() |>
  mutate(distFromTagSiteM = distVincentyEllipsoid(df.alltags.912.clean[,c("recvDeployLon", "recvDeployLat")],
                                                  df.alltags.912.clean[,c("tagDepLon", "tagDepLat")]),
         distFromTagSiteKM = distFromTagSiteM/1000,
         bandingCode = case_when(
           speciesEN == "Horned Lark" ~ "HOLA",
           speciesEN == "Chestnut-collared Longspur" ~ "CCLO",
           speciesEN == "Sprague's Pipit" ~ "SPPI",
           speciesEN == "Baird's Sparrow" ~ "BAIS",
           speciesEN == "Thick-billed Longspur" ~ "TBLO",
           speciesEN == "Grasshopper Sparrow" ~ "GRSP",
           TRUE ~ "NA"))

fallDepDates100km <- df.alltags.912.clean |>
  filter(distFromTagSiteKM <= 100 & tagDepYear == year & numDaysDetected > 1) |>
  group_by(motusTagID, speciesEN, bandingCode, tagDepYear, tagDepLat, tagDepLon, numDaysDetected) |>
  summarize(depDate100km = max(ts)) |>
  arrange(speciesEN, tagDepYear)

fallDepDates150km <- df.alltags.912.clean |>
  filter(distFromTagSiteKM <= 150 & tagDepYear == year & numDaysDetected > 1) |>
  group_by(motusTagID, speciesEN, bandingCode, tagDepYear, tagDepLat, tagDepLon, numDaysDetected) |>
  summarize(depDate150km = max(ts)) |>
  arrange(speciesEN, tagDepYear)

fallDepDates200km <- df.alltags.912.clean |>
  filter(distFromTagSiteKM <= 200 & tagDepYear == year & numDaysDetected > 1) |>
  group_by(motusTagID, speciesEN, bandingCode, tagDepYear, tagDepLat, tagDepLon, numDaysDetected) |>
  summarize(depDate200km = max(ts)) |>
  arrange(speciesEN, tagDepYear)

deps100 <- fallDepDates100km |>
  ungroup()|>
  select(motusTagID, depDate100km)

deps150 <- fallDepDates150km |>
  ungroup()|>
  select(motusTagID, depDate150km)

deps200 <- fallDepDates200km |>
  ungroup() |>
  select(motusTagID, depDate200km, speciesEN)

allDepDates <- deps200 |>
  left_join(deps150, by = "motusTagID") |>
  left_join(deps100, by = "motusTagID") |>
  select(motusTagID, speciesEN, depDate100km, depDate150km, depDate200km) |>
  arrange(speciesEN, depDate150km)

df.alltags.912.clean <- df.alltags.912.clean |>
  left_join(deps100, by = "motusTagID") 

df.alltags.912.clean <- df.alltags.912.clean |>
  left_join(deps150, by = "motusTagID")

df.alltags.912.clean <- df.alltags.912.clean |>
  left_join(deps200, by = "motusTagID") |>
  select(-speciesEN.y) |>
  rename(speciesEN = speciesEN.x)

# going to look at departures along with the next detection for each individual that has migratory detections
df_sorted <- df.alltags.912.clean %>%
  filter(!is.na(depDate100km) & ts >= depDate100km) |>
  arrange(motusTagID, recvDeployName, ts)

# pulling just the first post-departure detection
df_next_detection <- df_sorted |>
  filter(ts > depDate100km) |>
  group_by(motusTagID) |>
  arrange(ts) |>
  slice_head(n = 1)

# and just the departures themselves
df_deps <- df_sorted |>
  filter(ts == depDate100km)

# binding together
df_deps_next_det <- df_deps |>
  bind_rows(df_next_detection) |>
  select(motusTagID, speciesEN, ts, depDate100km, recvDeployName, recvDeployLat, distBtwRecvKM, timeD, numDaysDetected, tagDepYear) |>
  mutate(depMonth100 = month(depDate100km))|>
  arrange(speciesEN, motusTagID, depDate100km, ts)

migBirds <- df.alltags.912.clean |>
  filter(distFromTagSiteKM >100)

newDepDates <- df_deps_next_det |>
  mutate(typeOfDep = case_when(
    depMonth100 %in% c(8, 9, 10) & motusTagID %in% migBirds$motusTagID & timeD < 16 ~ "migDep", # departure month is reasonable and bird was detected not too long after departing
    depMonth100 %in% c(6, 7) & motusTagID %in% migBirds$motusTagID & !is.na(timeD) ~ "regionalMovement", # departure month is super early but they were detected again later on so the tag wasn't lost
    depMonth100 %in% c(6, 7) & !motusTagID %in% migBirds$motusTagID & is.na(timeD) ~ "possibleTagLoss", # departure month is super early and were not detected again
    depMonth100 %in% c(8, 9, 10) & motusTagID %in% migBirds$motusTagID & timeD > 16 ~ "migDepOrRegional", # departure month is realistic but the long gap between detections makes me less confident that the date is correct
    depMonth100 %in% c(8, 9, 10) & !motusTagID %in% migBirds$motusTagID ~ "possibleMigDep", # departure month is reasonable but were not detected again so can't be sure if they were migrating, tag dropped or regional movement
    TRUE ~ "unknown" # this would be everything else, aka anything with an early departure and that wasn't detected again - tag loss?
  )) |>
  group_by(motusTagID) |>
  filter(ts == max(ts)) |>
  select(-ts) |> 
  rename(nextRecv = recvDeployName,
         nextRecvLat = recvDeployLat, 
         distToNextRecvKM = distBtwRecvKM,
         timeToNextRecvDays = timeD) |>
  mutate(bandingCode = case_when(
    speciesEN == "Horned Lark" ~ "HOLA",
    speciesEN == "Chestnut-collared Longspur" ~ "CCLO",
    speciesEN == "Sprague's Pipit" ~ "SPPI",
    speciesEN == "Baird's Sparrow" ~ "BAIS",
    speciesEN == "Thick-billed Longspur" ~ "TBLO",
    speciesEN == "Grasshopper Sparrow" ~ "GRSP",
    TRUE ~ "NA"
  ))

```

The departure dates in the figure below are only those that I classified as a migratory departure or possible migratory departure. 

```{r fall departure date plot, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Fall departure dates, calculated as the last detection within 100km of the tagging location.", fig.align='left', fig.width=9, fig.height=6, fig.pos="H"}
# creating a month-day column just for plotting
newDepDates$Month_Day100 <- format(as.Date(newDepDates$depDate100km), "%m-%d")
newDepDates$Month_Day100 <- as.Date(newDepDates$Month_Day100, "%m-%d")

ggplot() +
  geom_boxplot(data = filter(newDepDates, typeOfDep %in% c("migDep", 
                                                           "possibleMigDep") & speciesEN %in% c("Sprague's Pipit", "Chestnut-collared Longspur")), aes(x = as.factor(tagDepYear), y = Month_Day100, fill = as.factor(speciesEN)), colour = "#505050", outlier.shape = NA, alpha = 0.3) +
  geom_point(data = filter(newDepDates, typeOfDep %in% c("migDep", 
                                                         "possibleMigDep")), aes(x = as.factor(tagDepYear), y = Month_Day100, fill = as.factor(speciesEN), shape = as.factor(typeOfDep)), size = 3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.1), colour = "black") +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%b %d")+
  coord_flip()+
  facet_wrap(~speciesEN) +
  theme_classic() +
  scale_colour_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e")) +
  scale_fill_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e")) +
  scale_shape_manual(values = c("migDep" = 21, "possibleMigDep" = 24), # Default shapes for each type
                     labels = c("Migratory departure", "Possible migratory departure")) +  # Custom labels for legend
  labs(y = "Departure Date", x = "Year") +
  guides(colour = "none", fill = "none", shape = guide_legend(title = "Type of Departure")) +
  theme(legend.position = "bottom",   # Place legend at the bottom
        legend.direction = "horizontal",  # Arrange legend items horizontally
        legend.justification = c("center", "top"),  # Center the legend horizontally
        legend.box = "horizontal",  # Ensure the items are in a horizontal box
        legend.box.margin = margin(t = 10)) +  # Optional: space above the legend +
  theme(text = element_text(size = 14)) +
   theme(text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

## Spring departure dates

We don't actually know for sure where the tagged GBMP birds overwintered, so these are more accurately described as the first date that each individual was detected moving north in the spring following tag deployment (one calendar year later than the year of deployment). I could only calculate this for birds with adequate spring migration detections, so some birds that were detected the following spring but only on the breeding grounds do not have spring "departure" dates.

```{r spring departure dates, echo = FALSE, warning = FALSE, message = FALSE}

springDepDates <- df.alltags.912.clean |>
  filter((year == tagDepYear + 1)) |>
  group_by(motusTagID, speciesEN) |>
  filter(recvDeployLat == min(recvDeployLat)) |>
  filter(ts == max(ts)) |>
  select(ts, motusTagID, bandingCode, recvDeployLat, recvDeployName) |>
  filter(recvDeployLat < 49) |>
  arrange(speciesEN, recvDeployLat)

```

```{r spring departure date plot, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Spring departure dates, calculated as the first detection moving northwards in the spring following the year of tag deployment.", fig.align='left', fig.width=8, fig.height=5, fig.pos="H"}

springDepDates$Month_Day <- format(as.Date(springDepDates$ts), "%m-%d") # creates a string
springDepDates$Month_Day <- as.Date(springDepDates$Month_Day, "%m-%d")

ggplot(springDepDates) +
  geom_point(aes(x = Month_Day, y = as.factor(year(ts)), fill = as.factor(speciesEN)), colour = "black", shape = 21, size = 4) +
  theme_classic() +
  scale_fill_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e")) +
  theme(text = element_text(size = 14)) +
  labs(x = "Date of first spring detection", y = "Year", fill = "Species")

```

# Migration pace

## Fall migration pace

This is a measurement of the average number of kilometers per day traveled by each individual between their departure location and their furthest south detection. 

```{r fall migration pace calculation, echo = FALSE, warning = FALSE, message = FALSE}

migBirds2 <- df.alltags.912.clean |>
  filter(distFromTagSiteKM > 200)

migPaceDets <- df.alltags.912.clean |>
  mutate(ordinalDate = yday(ts)) |>
  filter(!is.na(depDate100km) & ts >= depDate100km & year == tagDepYear & motusTagID %in% migBirds2$motusTagID) |>
  group_by(motusTagID, recvDeployName) |>
  filter(sig == max(sig)) # some birds have multiple detections with the same signal strength at an individual receiver

# now selecting just the departure date and southernmost point (for birds that have them)
migPaceDets2 <- migPaceDets |>
  group_by(motusTagID) |>
  arrange(ts) |>
  filter(ts == depDate100km | recvDeployLat == min(recvDeployLat)) |>
  select(motusTagID, speciesEN, tagDepYear, ts, recvDeployName, recvDeployLat, recvDeployLon, sig)

migPaceDets2 <- migPaceDets2 |>
  ungroup() |>
  arrange(motusTagID, ts) 

migPace1 <- migPaceDets2 |>
  mutate(timeDiff = if_else((motusTagID == lag(motusTagID) & recvDeployName != lag(recvDeployName )),
                               difftime(ts, lag(ts), units = "days"),
                               NA),
            totalDistM = if_else((motusTagID == lag(motusTagID) & recvDeployName != lag(recvDeployName)),
                                                 c(NA, distVincentyEllipsoid(migPaceDets2[,c("recvDeployLon", "recvDeployLat")])),
                                                 NA_real_),
            totalDistKM = totalDistM/1000,
            pace = totalDistKM/as.numeric(timeDiff)) |>
  filter(is.na(pace) == FALSE)

```

```{r fall migration pace plot, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Fall migration pace (km/day) of GBMP birds tagged in 2022, 2023 and 2024.", fig.align='left', fig.width=9, fig.height=4, fig.pos="H"}

ggplot(filter(migPace1, pace < 500)) +
  geom_point(aes(x = pace, y = as.factor(speciesEN), fill = as.factor(speciesEN), shape = as.factor(year(ts))), colour = "black", size = 4, position = position_jitter(height = 0.1, width = 0.1), alpha = 0.7) +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  labs(x = "Migration pace (km/day)", y = "Species", shape = "Year") +
  scale_colour_viridis_d() +
  scale_shape_manual(values = c(21, 22, 24)) +
  guides(fill = "none") +
  scale_fill_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e"))

```

Table 5. Average, maximum and minimum migration pace values (km/day) for GBMP tagged birds, by species.
```{r migration pace table, echo = FALSE, warning = FALSE, message = FALSE}

paceTable <- migPace1 |>
  filter(pace < 500) |>
  group_by(speciesEN) |>
  summarize(nTags = n_distinct(motusTagID),
            maxPace = max(pace),
            minPace = min(pace),
            avgPace = mean(pace)) |>
  mutate(across(where(is.numeric), ~ round(., 0)))

kable(paceTable,
      format = "latex",
      table.attr = "style='width:80%;'",
      col.names = c("Species", "Number of birds with pace values", "Highest pace (km/day)", "Lowest pace (km/day)", "Average pace (km/day)")
) %>% 
  kable_styling(position='left', latex_options = "scale_down") |>
  row_spec(0, bold = TRUE)

```

## Spring migration pace

```{r spring mig pace calculations, echo = FALSE, warning = FALSE, message = FALSE}

springArrivals <- df.alltags.912.clean |>
  filter(year == (tagDepYear + 1) & distFromTagSiteKM <= 100) |>
  group_by(motusTagID) |>
  arrange(ts) |>
  slice_head(n = 1) |>
  select(speciesEN, bandingCode, ts, recvDeployName, recvDeployLat, distFromTagSiteKM, tagDepYear) |>
  arrange(speciesEN, recvDeployLat) 

springArrivals$Month_Day <- format(as.Date(springArrivals$ts), "%m-%d") # creates a string
springArrivals$Month_Day <- as.Date(springArrivals$Month_Day, "%m-%d")

springArrivalDates <- springArrivals |>
  select(motusTagID, ts) |>
  rename(springArrivalDate = ts)

springMigPaceDets <- df.alltags.912.clean |>
  filter(year == (tagDepYear + 1))
   

springMigPaceDets <- springMigPaceDets |>
  left_join(springArrivalDates, by = "motusTagID") |>
  filter(is.na(springArrivalDate) == FALSE) |>
  group_by(motusTagID, recvDeployName) |>
  filter(sig == max(sig) | ts == springArrivalDate) |>
  select(motusTagID, speciesEN, tagDepYear, ts, recvDeployName, recvDeployLat, recvDeployLon, sig, springArrivalDate)

springMigPaceDets2 <- springMigPaceDets |>
  group_by(motusTagID) |>
  arrange(ts) |>
  filter(recvDeployLat == min(recvDeployLat) | ts == springArrivalDate) |>
  select(motusTagID, speciesEN, tagDepYear, ts, recvDeployName, recvDeployLat, recvDeployLon, sig, springArrivalDate)

springMigPaceDets2 <- springMigPaceDets2 |>
  ungroup() |>
  arrange(motusTagID, ts)

springMigPace1 <- springMigPaceDets2 |>
  mutate(timeDiff = if_else((motusTagID == lag(motusTagID) & recvDeployName != lag(recvDeployName )),
                            difftime(ts, lag(ts), units = "days"),
                            NA),
         totalDistM = if_else((motusTagID == lag(motusTagID) & recvDeployName != lag(recvDeployName)),
                              c(NA, distVincentyEllipsoid(springMigPaceDets2[,c("recvDeployLon", "recvDeployLat")])),
                              NA_real_),
         totalDistKM = totalDistM/1000,
         pace = totalDistKM/as.numeric(timeDiff)) |>
  filter(is.na(pace) == FALSE)

```


```{r spring mig pace plot, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Spring migration pace (km/day) of GBMP birds tagged in 2022 and 2023. Note that year refers to the spring following the year of tag deployment, thus there are no pace values for birds tagged in 2024 because they have not migrated back to the breeding grounds yet.", fig.align='left', fig.width=9, fig.height=4, fig.pos="H"}

ggplot(filter(springMigPace1, pace < 500)) +
  geom_point(aes(x = pace, y = as.factor(speciesEN), fill = as.factor(speciesEN), shape = as.factor(year(ts))), colour = "black", size = 4, position = position_jitter(height = 0.1, width = 0.1), alpha = 0.7) +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  labs(x = "Migration pace (km/day)", y = "Species", shape = "Year") +
  scale_shape_manual(values = c(21, 22, 24)) +
  guides(fill = "none") +
  scale_fill_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e"))

```
# Spring arrival dates

Fifteen of the birds tagged in 2022 and 2023 were detected returning to the breeding grounds in 2023 and 2024, respectively.

```{r spring arrival date plot, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Spring arrival dates, calculated as the first detection within 100km of the tagging location in the spring following the year of tag deployment.", fig.align='left', fig.width=8, fig.height=5, fig.pos="H"}

springArrivals$Month_Day <- format(as.Date(springArrivals$ts), "%m-%d") # creates a string
springArrivals$Month_Day <- as.Date(springArrivals$Month_Day, "%m-%d")

ggplot(springArrivals) +
  geom_point(aes(x = Month_Day, y = as.factor(year(ts)), fill = as.factor(speciesEN)), colour = "black", shape = 21, size = 4) +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  labs(x = "Spring breeding ground arrival date", y = "Year", fill = "Species") +
  scale_fill_manual(values = c("Sprague's Pipit" = "#35b779", "Horned Lark" = "#21918c", "Chestnut-collared Longspur" = "#443983", "Baird's Sparrow" = "#440154", "Thick-billed Longspur" = "#90d743", "Grasshopper Sparrow" = "#31688e"))


```

# Migration paths

The maps below show the apparent migration paths (fall and spring, for birds with detections in both seasons) for tagged GBMP birds. 

Most tagged birds seemed to head slightly east and then south, regardless of species or tagging location.

```{r migration paths by species, fig.cap="Apparent migration paths for the six GBMP grassland songbird species, as observed from Motus automated radio telemetry detections in 2022, 2023 and 2024. Purple shaded areas represent eBird nonbreeding ranges for each species.", echo = FALSE, message = FALSE, warning = FALSE, results = 'hide', fig.path = "figures/", fig.align='left', fig.width=10, fig.height=6, fig.pos="H"}

states_CanUS <- ne_states(country = c("canada", "united states of america"), returnclass = "sf")

# loading eBird vector data for winter ranges of each species
CCLO_range <- st_read("chclon_range_2022/chclon_range_2022.gpkg")
CCLO_winter_range <- CCLO_range |>
  filter(season == "nonbreeding")

SPPI_range <- st_read("sprpip_range_2022/sprpip_range_2022.gpkg")
SPPI_winter_range <- SPPI_range |>
  filter(season == "nonbreeding")

BAIS_range <- st_read("baispa_range_2022/baispa_range_2022.gpkg")
BAIS_winter_range <- BAIS_range |>
  filter(season == "nonbreeding")

HOLA_range <- st_read("horlar_range_2022/horlar_range_2022.gpkg")
HOLA_winter_range <- HOLA_range |>
  filter(season == "nonbreeding")

TBLO_range <- st_read("mcclon_range_2022/mcclon_range_2022.gpkg")
TBLO_winter_range <- TBLO_range |>
  filter(season == "nonbreeding")

GRSP_range <- st_read("graspa_range_2022/graspa_range_2022.gpkg")
GRSP_winter_range <- GRSP_range |>
  filter(season == "nonbreeding")

allWinterRanges <- CCLO_winter_range |>
  bind_rows(SPPI_winter_range, BAIS_winter_range, HOLA_winter_range, TBLO_winter_range, GRSP_winter_range) |>
  rename(speciesEN = common_name)
  
fun.getpath <- function(df) {
  df %>%
    filter(!is.na(recvDeployLat) | !(recvDeployLat == 0)) %>% 
    group_by(motusTagID, mfgID, runID, recvDeployName, ambigID, 
             tagDepLon, tagDepLat, recvDeployLat, recvDeployLon, speciesEN, year) %>%
    summarize(max.runLen = max(runLen), time = mean(ts), .groups = "drop") %>%
    arrange(motusTagID, time) %>%
    data.frame()
}

path912 <- fun.getpath(df.alltags.912.clean)

tagDepYears <- tags912 |>
  select(motusTagID, utcYearStart) |>
  rename(tagDepYear = utcYearStart) |>
  filter(is.na(tagDepYear) == FALSE)

path912 <- path912 |>
  left_join(tagDepYears, by = "motusTagID")

colnames(path912)

first_receivers <- path912 %>%
  group_by(motusTagID) %>%
  slice(1) %>%
  ungroup()

path912$tagSite <- "Tag Site"
activeRecv$recvType <- "Active Receivers"
allWinterRanges$Season <- "Nonbreeding Season Range"
path912$recvWithDets <- "Receiver With Detections"
path912$lineType <- "Apparent Migration Route"
first_receivers$lineType <- "Apparent Migration Route"
path912 <- path912 |>
  mutate(migrationSeason = case_when(
    as.numeric(year(time)) == tagDepYear ~ "Apparent Fall Migration Route",
    as.numeric(year(time)) == (tagDepYear + 1) ~ "Apparent Spring Migration Route",
    TRUE ~ "NA"
  ))

xmin <- min(path912$recvDeployLon, na.rm = TRUE) - 2
xmax <- max(path912$recvDeployLon, na.rm = TRUE) + 10
ymin <- min(path912$recvDeployLat, na.rm = TRUE) - 1
ymax <- max(path912$recvDeployLat, na.rm = TRUE) + 1

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = allWinterRanges, aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "") +
  geom_path(data = filter(path912, is.na(speciesEN) == FALSE), 
            aes(x = recvDeployLon, y = recvDeployLat, 
                colour = as.factor(speciesEN), group = motusTagID), show.legend = FALSE) +
  geom_segment(data = first_receivers, aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat, colour = speciesEN), show.legend = FALSE) +
  geom_point(data = filter(path912, is.na(speciesEN) == FALSE ), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, is.na(speciesEN) == FALSE), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  facet_wrap(~speciesEN) +
  # Manual color scales for different layers
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
  scale_colour_manual(values = c("Baird's Sparrow" = "#440154", "Chestnut-collared Longspur" = "#443983", "Grasshopper Sparrow" = "#31688e", "Horned Lark" = "#21918c", "Sprague's Pipit" = "#35b779", "Thick-billed Longspur" = "#90d743")) +
  
  # Create a combined, unified legend using guides()
  guides(
    fill = guide_legend(order = 2, title = "Legend", ncol = 1))
```

## Baird\'s Sparrow

One Baird's Sparrow was detected in a possible overwintering location, while others were detected during their fall migration heading south-east into Montana, South Dakota and Colorado.

```{r BAIS individual maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=5, fig.height=4, fig.align='left', fig.cap="Apparent migration paths for two Baird's Sparrows that were detected in possible overwintering locations.", fig.pos="H"}

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Baird's Sparrow"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "BAIS-67391") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 67391), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#440154") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 67391), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#440154") +
 geom_segment(data = filter(first_receivers, motusTagID == 67391), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#440154") +
  geom_point(data = filter(path912, motusTagID == 67391), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 67391), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(breaks = c(-110, -100, -90, -80)) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Baird's Sparrow"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "BAIS-85292") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 85292), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#440154") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 85292), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#440154") +
 geom_segment(data = filter(first_receivers, motusTagID == 85292), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#440154") +
  geom_point(data = filter(path912, motusTagID == 85292), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 85292), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

```

## Chestnut-collared Longspur

Two Chestnut-collared Longspur were detected in New Mexico, which is within the known wintering range of the species (Bleho et al. 2020). Note that CCLO-67387 was detected back on the breeding grounds in the spring, however it had no other spring detections so the tracks do not show up on the map below. The coloured area on the map represents the eBird CCLO non-breeding range.

```{r CCLO individual maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=5, fig.height=4, fig.align='left', fig.cap="Apparent migration paths for two Chestnut-collared Longspurs that were detected in possible overwintering locations.", fig.pos="H"}

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Chestnut-collared Longspur"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "CCLO-85818") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 85818), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#443983") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 85818), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#443983") +
 geom_segment(data = filter(first_receivers, motusTagID == 85818), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#443983") +
  geom_point(data = filter(path912, motusTagID == 85818), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 85818), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(breaks = c(-110, -100, -90, -80)) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Chestnut-collared Longspur"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "CCLO-67387") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 67387), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#443983") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 67387), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#443983") +
 geom_segment(data = filter(first_receivers, motusTagID == 67387), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#443983") +
  geom_point(data = filter(path912, motusTagID == 67387), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 67387), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

```

## Grasshopper Sparrow

One Grasshopper Sparrow was detected heading south-east into Montana during fall migration.

```{r GRSP individual maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=7, fig.height=5, fig.align='left', fig.cap="Apparent migration path for one Grasshopper Sparrow.", fig.pos="H"}

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Grasshopper Sparrow"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "GRSP-85288") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 85288), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#31688e") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 85288), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#31688e") +
 geom_segment(data = filter(first_receivers, motusTagID == 85288), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#31688e") +
  geom_point(data = filter(path912, motusTagID == 85288), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 85288), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(breaks = c(-110, -100, -90, -80)) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

```

## Horned Lark

Only three of 15 detected Horned Larks had migratory detections. The year-round range for Horned Lark includes most of the USA (Beason et al. 2020), so the last detections for our three tagged birds may be in plausible overwintering locations. The coloured area on the map represents the eBird HOLA non-breeding range.

```{r HOLA individual maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=5, fig.height=4, fig.align='left', fig.cap="Apparent migration paths for two Horned Larks.", fig.pos="H"}

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Horned Lark"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "HOLA-75828") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 75828), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#21918c") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 75828), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#21918c") +
 geom_segment(data = filter(first_receivers, motusTagID == 75828), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#21918c") +
  geom_point(data = filter(path912, motusTagID == 75828), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 75828), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Horned Lark"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "HOLA-75830") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 75830), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#21918c") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 75830), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#21918c") +
 geom_segment(data = filter(first_receivers, motusTagID == 75830), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#21918c") +
  geom_point(data = filter(path912, motusTagID == 75830), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 75830), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

```

## Sprague\'s Pipit

One Sprague's Pipit was detected in Texas and another one was detected in northern Mexico, both of which are within the known wintering range of the species (Davis et al. 2020). Migration tracks for many tagged pipits ended in Kansas or Oklahoma, which seems like an unlikely endpoint considering that it is north of known wintering areas and eBird detections for Sprague's Pipit are minimal in those areas during winter months (Fink et al. 2023). The coloured area on the map represents the eBird SPPI non-breeding range.

```{r SPPI individual maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=5, fig.height=4, fig.align='left', fig.cap="Apparent migration paths for four Sprague's Pipits that were detected in possible overwintering locations.", fig.pos="H"}

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Sprague's Pipit"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "SPPI-61024") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 61024), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#35b779") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 61024), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#35b779") +
 geom_segment(data = filter(first_receivers, motusTagID == 61024), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#35b779") +
  geom_point(data = filter(path912, motusTagID == 61024), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 61024), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Sprague's Pipit"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "SPPI-75413") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 75413), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#35b779") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 75413), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#35b779") +
 geom_segment(data = filter(first_receivers, motusTagID == 75413), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#35b779") +
  geom_point(data = filter(path912, motusTagID == 75413), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 75413), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))


ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Sprague's Pipit"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "SPPI-61020") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 61020), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#35b779") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 61020), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#35b779") +
 geom_segment(data = filter(first_receivers, motusTagID == 61020), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#35b779") +
  geom_point(data = filter(path912, motusTagID == 61020), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 61020), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))


ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Sprague's Pipit"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "SPPI-75837") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 75837), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#35b779") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 75837), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#35b779") +
 geom_segment(data = filter(first_receivers, motusTagID == 75837), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#35b779") +
  geom_point(data = filter(path912, motusTagID == 75837), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 75837), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))


```

## Thick-billed Longspur

Three Thick-billed Longspur were detected in possible overwintering locations in Colorado (tracks for two are shown below).

```{r TBLO individual maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=5, fig.height=4, fig.align='left', fig.cap="Apparent migration paths for two Thick-billed Longspur that were detected in possible overwintering locations.", fig.pos="H"}

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Thick-billed Longspur"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "TBLO-74693") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 74693), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#90d743") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 74693), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#90d743") +
 geom_segment(data = filter(first_receivers, motusTagID == 74693), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#90d743") +
  geom_point(data = filter(path912, motusTagID == 74693), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 74693), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = filter(allWinterRanges, speciesEN == "Thick-billed Longspur"), aes(fill = as.factor(Season)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  labs(x = "", y = "", title = "TBLO-86990") +
    geom_point(data = activeRecv, aes(x = recvDeployLon, y = recvDeployLat, colour = as.factor(recvType)), shape = 1) +
  geom_path(data = filter(path912, as.numeric(year(time)) == tagDepYear & motusTagID == 86990), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#90d743") +
    geom_path(data = filter(path912, as.numeric(year(time)) == (tagDepYear + 1) & motusTagID == 86990), 
              aes(x = recvDeployLon, y = recvDeployLat, group = motusTagID, linetype = as.factor(migrationSeason)), size = 1, colour = "#90d743") +
 geom_segment(data = filter(first_receivers, motusTagID == 86990), aes(x = tagDepLon, y = tagDepLat, xend = recvDeployLon, yend = recvDeployLat), show.legend = FALSE, size = 1, colour = "#90d743") +
  geom_point(data = filter(path912, motusTagID == 86990), aes(x = recvDeployLon, y = recvDeployLat, fill = as.factor(recvWithDets)), shape = 21) + # plotting receivers
  geom_point(data = filter(path912, motusTagID == 86990), # plotting tag deployment sites
             aes(x = tagDepLon, y = tagDepLat, fill = as.factor(tagSite)), colour = "black", shape = 23, size = 2) +
  scale_x_continuous(
    breaks = c(-110, -100, -90, -80)
  ) +
  geom_text(data = filter(states_CanUS, postal %in% c("MT", "ND", "WY", "SD", "NE", "KS", "OK", "NM", "TX", "CO", "WY", "FL", "ID", "UT", "AZ")), aes(x = longitude, y = latitude, label = postal)) +
      geom_text(data = filter(states_CanUS, postal %in% c("AB", "SK", "MB")), aes(label = postal), y = 51, x = c(-112.5, -105, -97.5)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "Tag Site" = "red", "Receiver With Detections" = "black")) +
    scale_colour_manual(values = c("Active Receivers" = "gray45")) +
    scale_linetype_manual(values = c("Apparent Fall Migration Route" = "solid", "Apparent Spring Migration Route" = "dotdash")) +
    
    # Create a combined, unified legend using guides()
    guides(
      color = guide_legend(order = 1, title = "Legend", ncol = 1),
      fill = guide_legend(order = 2, title = NULL, ncol = 1),
      shape = guide_legend(order = 3, ncol = 1),
      linetype = guide_legend(order = 4, title = NULL, ncol = 1)
    ) +  # This ensures that shapes are displayed in the legend
    theme(legend.spacing.y = unit(0, "cm"))

```

# Overwintering Areas and Stopover Sites

As shown on the migration route maps above, only a few tagged birds had detections within known overwintering ranges. However, most of those detections were transitory in nature, spanning only a few minutes or so. To identify possible stopover sites or overwintering areas, I pulled any post-departure detections that spanned at least 24h either at a single receiver or a group of receivers within about 50km of one another. I would consider these possible stopover sites for detections outside of the wintering range, whereas any "stopovers" within the known wintering range of each species may represent wintering sites. Date ranges for each "stopover" are shown on each map.

```{r CCLO stopover prep code, echo = FALSE, message = FALSE, warning = FALSE }

CCLO <- df.alltags.912.clean %>%
  filter(speciesEN == "Chestnut-collared Longspur" & ts > depDate100km)

# Function to calculate distance between two points (Vincenty formula)
calculate_distance <- function(lat1, lon1, lat2, lon2) {
  distVincentyEllipsoid(c(lon1, lat1), c(lon2, lat2))} # Returns distance in meters

# Function to calculate distance between two points (Vincenty formula)
calculate_distance <- function(lat1, lon1, lat2, lon2) {
  distVincentyEllipsoid(c(lon1, lat1), c(lon2, lat2))} # Returns distance in meters

# create an empty list to store detections in
valid_detections <- list()

for(tag_id in unique(CCLO$motusTagID)) {
  
  # Filter data for the current bird (motusTagID)
  bird_data <- filter(CCLO, motusTagID == tag_id) %>%
    group_by(runID) %>%
    slice_head(n = 1) %>%  # just keeping the first hit in each run to reduce the amount of data I'm using
    ungroup() %>%
    group_by(motusTagID, recvDeployName, ts) %>%
    filter(sig == max(sig)) %>%
    ungroup() %>%
    arrange(ts)
  
  # Create an empty list to track groupings
  groupings <- list()
  
  assigned_detections <- c()
  
  # Loop through each detection and check its distance from others
  for(i in 1:nrow(bird_data)) {
    # Skip if this detection has already been assigned to a group
    if (i %in% assigned_detections) next
    
    current_row <- bird_data[i, ]
    
    # Initialize a group with the current detection
    group <- list(current_row)
    
    # Check if the current detection is within 50km of any other detection
    for(j in 1:nrow(bird_data)) {
      if(i != j && !(j %in% assigned_detections)) { # each run should only be used in a single group
        other_row <- bird_data[j, ]
        
        # Calculate the distance between the two detections if the receiver names are different
        if (current_row$year == other_row$year) {  # current_row$recvDeployName != other_row$recvDeployName & 
          dist <- calculate_distance(
            current_row$recvDeployLat, current_row$recvDeployLon,
            other_row$recvDeployLat, other_row$recvDeployLon
          )
          
          # If within 70 km, add to the same group
          if(dist <= 70000) {
            group <- append(group, list(other_row))
            assigned_detections <- c(assigned_detections, j)  # Mark detection as assigned
          }
        }
      }
    }
    
    # Only keep groups that have detections spanning at least 24 hours
    group_df <- do.call(rbind, group)
    min_ts <- min(group_df$ts)
    max_ts <- max(group_df$ts)
    
    # Calculate the time difference between the first and last detection
    time_diff_hours <- as.numeric(difftime(max_ts, min_ts, units = "hours"))
    
    # If time span is at least 24 hours, store the group (can adjust to any desired timespan)
    if(time_diff_hours >= 24) {
      # Create a unique group identifier (can use a combination of tag_id and min_ts)
      group_id <- paste(tag_id, min_ts, sep = "-")
      
      # Add the group_id to each detection in the group
      group_df$group_id <- group_id
      
      # Store the group with the unique group_id
      groupings[[group_id]] <- group_df
    }
  }
  
  # After grouping, add valid groupings to final results
  for(group_name in names(groupings)) {
    # If valid_detections is a data frame, we need to bind the results properly
    if (exists("valid_detections")) {
      valid_detections <- rbind(valid_detections, groupings[[group_name]])
    } else {
      # If it's the first iteration, initialize the valid_detections data frame
      valid_detections <- groupings[[group_name]]
    }
  }
}

# Combine the valid detections into one data frame
valid_detections_df <- bind_rows(valid_detections) %>%
  group_by(motusTagID, group_id) %>%
  arrange(motusTagID, group_id, ts)

# Calculate time difference from the first detection for each motusTagID
valid_detections_df <- valid_detections_df %>%
  group_by(group_id) %>%
  mutate(
    first_detection_time = first(ts),
    time_diff_hours = as.numeric(difftime(ts, first_detection_time, units = "hours"))
  )

# Calculate distance to previous receiver only if the receiver names are different
valid_detections_df <- valid_detections_df %>%
  ungroup() %>%
  arrange(motusTagID, ts) %>%
  mutate(
    distance_from_previous_recv = c(NA, sapply(2:nrow(valid_detections_df), function(i) {
      if(valid_detections_df$motusTagID[i] == valid_detections_df$motusTagID[i - 1] &
         valid_detections_df$recvDeployName[i] != valid_detections_df$recvDeployName[i - 1] ) {
        calculate_distance(
          valid_detections_df$recvDeployLat[i], valid_detections_df$recvDeployLon[i],
          valid_detections_df$recvDeployLat[i - 1], valid_detections_df$recvDeployLon[i - 1]
        )
      } else {
        NA  # No distance calculation if receiver names are the same
      }
    }))
  )

valid_detections_df <- valid_detections_df |>
  dplyr::select(ts, runID, motusTagID, tagDepLat, tagDepLon, group_id, recvDeployLat, recvDeployLon, recvDeployName, speciesEN, year, daysSinceDep, date, depDate100km, distBtwRecvKM, distance_from_previous_recv, time_diff_hours, timeDiff)

# save each one as a different object so that I can re-run the code for the other species without losing the data
valid_detections_df_CCLO <- valid_detections_df

plotDataCCLO <- valid_detections_df_CCLO |>
  group_by(group_id) |>
  mutate(stopoverLengthDays = difftime(max(ts), min(ts), units = "days"),
         stopoverLengthRound = round(stopoverLengthDays, digits = 0),
         meanLat = mean(recvDeployLat),
         meanLon = mean(recvDeployLon),
         min_ts = as.Date(min(ts)), # adding the start and end points for plotting
         max_ts = as.Date(max(ts))) |>
  filter(recvDeployLat < 49) |>
  mutate(siteType = "Tag site")

```

```{r CCLO stopover maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=10, fig.height=8, fig.align='left', fig.cap="Possible stopover and overwintering locations for Chestnut-collared Longspur.", fig.pos="H"}

xmin <- min(plotDataCCLO$meanLon, na.rm = TRUE) - 10
xmax <- max(plotDataCCLO$meanLon, na.rm = TRUE) + 10
ymin <- min(plotDataCCLO$meanLat, na.rm = TRUE) - 5
ymax <- max(plotDataCCLO$meanLat, na.rm = TRUE) + 8

CCLO_range <- st_read("chclon_range_2022/chclon_range_2022.gpkg")
CCLO_winter_range <- CCLO_range |>
  filter(season == "nonbreeding") |>
  mutate(season2 = "Nonbreeding Season Range")

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = CCLO_winter_range, aes(fill = as.factor(season2)), color = NA, alpha = 0.3) + 
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  geom_point(data = plotDataCCLO, aes(x = tagDepLon, y = tagDepLat), fill = "red", colour = "black", shape = 23) +
  geom_point(data = filter(plotDataCCLO, !motusTagID == 61596), aes(x = meanLon, y = meanLat, size = as.numeric(stopoverLengthRound), fill = as.factor(motusTagID)), shape = 21) +
  geom_point(data = filter(plotDataCCLO, motusTagID == 61596), aes(x = meanLon, y = meanLat, size = as.numeric(stopoverLengthRound), fill = as.factor(motusTagID)), shape = 21) +
  labs(x = "", y = "") +
  geom_text(data = states_CanUS, aes(x = longitude, y = latitude, label = postal)) + 
  geom_label(data = filter(plotDataCCLO, motusTagID == 61596), aes(y = tagDepLat), x = -108.5, label = "Tag Sites", colour = "black", size = 3, hjust = 0.75, vjust = -0.5) +
  geom_label(data = filter(plotDataCCLO, motusTagID == 61587), aes(x = meanLon, label = paste("Date range:", min_ts, "-", max_ts)), color = "black", size = 3, vjust = -1, y = 39) + 
  geom_segment(data = filter(plotDataCCLO, motusTagID == 61587), aes(x = meanLon, y = 38.7, 
                                                                     xend = meanLon, yend = 39.5),
               color = "black", size = 0.5, linetype = "solid", arrow = arrow(type = "open", length = unit(0.05, "inches"))) +
  geom_label(data = filter(plotDataCCLO, motusTagID == 67508), aes(y = meanLat, label = paste("Date range:", min_ts, "-", max_ts)), color = "black", size = 3, vjust = -1, x = -100) +
  geom_segment(data = filter(plotDataCCLO, motusTagID == 67508), x = -102, y = 48.6, 
                                                                     xend = -107.5, yend = 48.1,
               color = "black", size = 0.5, linetype = "solid", arrow = arrow(type = "open", length = unit(0.05, "inches"))) +
  geom_label(data = filter(plotDataCCLO, motusTagID == 61596), aes(x = meanLon, label = paste("Date range:", min_ts, "-", max_ts)), color = "black", size = 3, vjust = -1, y = 45) +
  geom_segment(data = filter(plotDataCCLO, motusTagID == 61596), aes(x = meanLon, y = 46.75, 
                                                                     xend = meanLon, yend = 47.6),
               color = "black", size = 0.5, linetype = "solid", arrow = arrow(type = "open", length = unit(0.05, "inches"))) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  # Adjust the legend for stopoverLengthDays to range from 17 (shortest stopover) to 25 (longest stopover) by 2
  scale_size_continuous(name = "Stopover Length (days)", 
                        breaks = seq(min(as.numeric(plotDataCCLO$stopoverLengthRound)), max(as.numeric(plotDataCCLO$stopoverLengthRound)), by = 2),  # Define the breaks for the legend
                        labels = seq(min(as.numeric(plotDataCCLO$stopoverLengthRound)), max(as.numeric(plotDataCCLO$stopoverLengthRound)), by = 2),  # Define the labels for the legend
                        range = c(2, 10)) +
  scale_fill_manual(values = c("Nonbreeding Season Range" = "purple4", "61587" = "#440154", "67508" = "#21918c", "61596" = "#fde725")) +
  guides(
    fill = guide_legend(order = 2, title = "Legend", ncol = 1))


```

```{r HOLA stopover prep code, echo = FALSE, message = FALSE, warning = FALSE }

HOLA <- df.alltags.912.clean %>%
  filter(speciesEN == "Horned Lark" & ts > depDate100km)

# Function to calculate distance between two points (Vincenty formula)
calculate_distance <- function(lat1, lon1, lat2, lon2) {
  distVincentyEllipsoid(c(lon1, lat1), c(lon2, lat2))} # Returns distance in meters

# create an empty list to store detections in
valid_detections <- list()

for(tag_id in unique(HOLA$motusTagID)) {
  
  # Filter data for the current bird (motusTagID)
  bird_data <- filter(HOLA, motusTagID == tag_id) %>%
    group_by(runID) %>%
    slice_head(n = 1) %>%  # just keeping the first hit in each run to reduce the amount of data I'm using
    ungroup() %>%
    group_by(motusTagID, recvDeployName, ts) %>%
    filter(sig == max(sig)) %>%
    ungroup() %>%
    arrange(ts)
  
  # Create an empty list to track groupings
  groupings <- list()
  
  assigned_detections <- c()
  
  # Loop through each detection and check its distance from others
  for(i in 1:nrow(bird_data)) {
    # Skip if this detection has already been assigned to a group
    if (i %in% assigned_detections) next
    
    current_row <- bird_data[i, ]
    
    # Initialize a group with the current detection
    group <- list(current_row)
    
    # Check if the current detection is within 50km of any other detection
    for(j in 1:nrow(bird_data)) {
      if(i != j && !(j %in% assigned_detections)) { # each run should only be used in a single group
        other_row <- bird_data[j, ]
        
        # Calculate the distance between the two detections if the receiver names are different
        if (current_row$year == other_row$year) {  # current_row$recvDeployName != other_row$recvDeployName & 
          dist <- calculate_distance(
            current_row$recvDeployLat, current_row$recvDeployLon,
            other_row$recvDeployLat, other_row$recvDeployLon
          )
          
          # If within 70 km, add to the same group
          if(dist <= 70000) {
            group <- append(group, list(other_row))
            assigned_detections <- c(assigned_detections, j)  # Mark detection as assigned
          }
        }
      }
    }
    
    # Only keep groups that have detections spanning at least 24 hours
    group_df <- do.call(rbind, group)
    min_ts <- min(group_df$ts)
    max_ts <- max(group_df$ts)
    
    # Calculate the time difference between the first and last detection
    time_diff_hours <- as.numeric(difftime(max_ts, min_ts, units = "hours"))
    
    # If time span is at least 24 hours, store the group (can adjust to any desired timespan)
    if(time_diff_hours >= 24) {
      # Create a unique group identifier (can use a combination of tag_id and min_ts)
      group_id <- paste(tag_id, min_ts, sep = "-")
      
      # Add the group_id to each detection in the group
      group_df$group_id <- group_id
      
      # Store the group with the unique group_id
      groupings[[group_id]] <- group_df
    }
  }
  
  # After grouping, add valid groupings to final results
  for(group_name in names(groupings)) {
    # If valid_detections is a data frame, we need to bind the results properly
    if (exists("valid_detections")) {
      valid_detections <- rbind(valid_detections, groupings[[group_name]])
    } else {
      # If it's the first iteration, initialize the valid_detections data frame
      valid_detections <- groupings[[group_name]]
    }
  }
}

# Combine the valid detections into one data frame
valid_detections_df <- bind_rows(valid_detections) %>%
  group_by(motusTagID, group_id) %>%
  arrange(motusTagID, group_id, ts)

# Calculate time difference from the first detection for each motusTagID
valid_detections_df <- valid_detections_df %>%
  group_by(group_id) %>%
  mutate(
    first_detection_time = first(ts),
    time_diff_hours = as.numeric(difftime(ts, first_detection_time, units = "hours"))
  )

# Calculate distance to previous receiver only if the receiver names are different
valid_detections_df <- valid_detections_df %>%
  ungroup() %>%
  arrange(motusTagID, ts) %>%
  mutate(
    distance_from_previous_recv = c(NA, sapply(2:nrow(valid_detections_df), function(i) {
      if(valid_detections_df$motusTagID[i] == valid_detections_df$motusTagID[i - 1] &
         valid_detections_df$recvDeployName[i] != valid_detections_df$recvDeployName[i - 1] ) {
        calculate_distance(
          valid_detections_df$recvDeployLat[i], valid_detections_df$recvDeployLon[i],
          valid_detections_df$recvDeployLat[i - 1], valid_detections_df$recvDeployLon[i - 1]
        )
      } else {
        NA  # No distance calculation if receiver names are the same
      }
    }))
  )

valid_detections_df_HOLA <- valid_detections_df |>
  dplyr::select(ts, runID, motusTagID, tagDepLat, tagDepLon, group_id, recvDeployLat, recvDeployLon, recvDeployName, speciesEN, year, daysSinceDep, date, depDate100km, distBtwRecvKM, distance_from_previous_recv, time_diff_hours, timeDiff)

plotDataHOLA <- valid_detections_df_HOLA |>
  group_by(group_id) |>
  mutate(stopoverLengthDays = difftime(max(ts), min(ts), units = "days"),
         stopoverLengthRound = round(stopoverLengthDays, digits = 0),
         meanLat = mean(recvDeployLat),
         meanLon = mean(recvDeployLon),
         min_ts = as.Date(min(ts)), # adding the start and end points for plotting
         max_ts = as.Date(max(ts))) |>
  filter(recvDeployLat < 49) |>
  mutate(siteType = "Tag site")

```

```{r HOLA stopover maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=10, fig.height=8, fig.align='left', fig.cap="Possible stopover and overwintering locations for Horned Lark.", fig.pos="H"}

xmin <- min(plotDataHOLA$meanLon, na.rm = TRUE) - 10
xmax <- max(plotDataHOLA$meanLon, na.rm = TRUE) + 10
ymin <- min(plotDataHOLA$meanLat, na.rm = TRUE) - 5
ymax <- max(plotDataHOLA$meanLat, na.rm = TRUE) + 10

HOLA_range <- st_read("horlar_range_2022/horlar_range_2022.gpkg")
HOLA_winter_range <- HOLA_range |>
  filter(season == "nonbreeding") |>
  mutate(season2 = "Nonbreeding season range")

ggplot(data = world) + 
  geom_sf(colour = "black") +
  geom_sf(data = HOLA_winter_range, color = NA, alpha = 0.3, aes(fill = as.factor(season2))) +
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  geom_point(data = plotDataHOLA, aes(x = tagDepLon, y = tagDepLat), colour = "black", fill = "red", shape = 23) +
  geom_point(data = plotDataHOLA, aes(x = meanLon, y = meanLat, size = as.numeric(stopoverLengthRound), fill = as.factor(motusTagID)), shape = 21) +
  labs(x = "", y = "") +
  geom_text(data = states_CanUS, aes(x = longitude, y = latitude, label = postal)) + 
  geom_label(data = plotDataHOLA, aes(x = tagDepLon, y = tagDepLat), label = "Tag Site", colour = "black", size = 3, hjust = 0.5, vjust = -0.5) +
  geom_label(data = filter(plotDataHOLA, motusTagID == 75828), aes(label = paste("Date range:", min_ts, "-", max_ts)), color = "black", size = 3, vjust = -1, y = 41.5, x = -113) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  # Adjust the legend for stopoverLengthDays to range from 17 (shortest stopover) to 25 (longest stopover) by 2
  scale_size_continuous(name = "Stopover Length (days)", 
                        breaks = seq(min(as.numeric(plotDataHOLA$stopoverLengthRound)), max(as.numeric(plotDataHOLA$stopoverLengthRound)), by = 2),  # Define the breaks for the legend
                        labels = seq(min(as.numeric(plotDataHOLA$stopoverLengthRound)), max(as.numeric(plotDataHOLA$stopoverLengthRound)), by = 2),  # Define the labels for the legend
                        range = c(2, 10)) +  # Set the size range
  scale_fill_manual(values = c("Nonbreeding season range" = "purple4", "75828" = "#21918c")) +
  guides(fill = guide_legend(order = 2, title = "Legend", ncol = 1))

```

```{r SPPI stopover prep code, echo = FALSE, message = FALSE, warning = FALSE }

SPPI <- df.alltags.912.clean %>%
  filter(speciesEN == "Sprague's Pipit" & ts > depDate100km)

# Function to calculate distance between two points (Vincenty formula)
calculate_distance <- function(lat1, lon1, lat2, lon2) {
  distVincentyEllipsoid(c(lon1, lat1), c(lon2, lat2))} # Returns distance in meters

# create an empty list to store detections in
valid_detections <- list()

for(tag_id in unique(SPPI$motusTagID)) {
  
  # Filter data for the current bird (motusTagID)
  bird_data <- filter(SPPI, motusTagID == tag_id) %>%
    group_by(runID) %>%
    slice_head(n = 1) %>%  # just keeping the first hit in each run to reduce the amount of data I'm using
    ungroup() %>%
    group_by(motusTagID, recvDeployName, ts) %>%
    filter(sig == max(sig)) %>%
    ungroup() %>%
    arrange(ts)
  
  # Create an empty list to track groupings
  groupings <- list()
  
  assigned_detections <- c()
  
  # Loop through each detection and check its distance from others
  for(i in 1:nrow(bird_data)) {
    # Skip if this detection has already been assigned to a group
    if (i %in% assigned_detections) next
    
    current_row <- bird_data[i, ]
    
    # Initialize a group with the current detection
    group <- list(current_row)
    
    # Check if the current detection is within 50km of any other detection
    for(j in 1:nrow(bird_data)) {
      if(i != j && !(j %in% assigned_detections)) { # each run should only be used in a single group
        other_row <- bird_data[j, ]
        
        # Calculate the distance between the two detections if the receiver names are different
        if (current_row$year == other_row$year) {  # current_row$recvDeployName != other_row$recvDeployName & 
          dist <- calculate_distance(
            current_row$recvDeployLat, current_row$recvDeployLon,
            other_row$recvDeployLat, other_row$recvDeployLon
          )
          
          # If within 70 km, add to the same group
          if(dist <= 70000) {
            group <- append(group, list(other_row))
            assigned_detections <- c(assigned_detections, j)  # Mark detection as assigned
          }
        }
      }
    }
    
    # Only keep groups that have detections spanning at least 24 hours
    group_df <- do.call(rbind, group)
    min_ts <- min(group_df$ts)
    max_ts <- max(group_df$ts)
    
    # Calculate the time difference between the first and last detection
    time_diff_hours <- as.numeric(difftime(max_ts, min_ts, units = "hours"))
    
    # If time span is at least 24 hours, store the group (can adjust to any desired timespan)
    if(time_diff_hours >= 24) {
      # Create a unique group identifier (can use a combination of tag_id and min_ts)
      group_id <- paste(tag_id, min_ts, sep = "-")
      
      # Add the group_id to each detection in the group
      group_df$group_id <- group_id
      
      # Store the group with the unique group_id
      groupings[[group_id]] <- group_df
    }
  }
  
  # After grouping, add valid groupings to final results
  for(group_name in names(groupings)) {
    # If valid_detections is a data frame, we need to bind the results properly
    if (exists("valid_detections")) {
      valid_detections <- rbind(valid_detections, groupings[[group_name]])
    } else {
      # If it's the first iteration, initialize the valid_detections data frame
      valid_detections <- groupings[[group_name]]
    }
  }
}

# Combine the valid detections into one data frame
valid_detections_df <- bind_rows(valid_detections) %>%
  group_by(motusTagID, group_id) %>%
  arrange(motusTagID, group_id, ts)

# Calculate time difference from the first detection for each motusTagID
valid_detections_df <- valid_detections_df %>%
  group_by(group_id) %>%
  mutate(
    first_detection_time = first(ts),
    time_diff_hours = as.numeric(difftime(ts, first_detection_time, units = "hours"))
  )

# Calculate distance to previous receiver only if the receiver names are different
valid_detections_df <- valid_detections_df %>%
  ungroup() %>%
  arrange(motusTagID, ts) %>%
  mutate(
    distance_from_previous_recv = c(NA, sapply(2:nrow(valid_detections_df), function(i) {
      if(valid_detections_df$motusTagID[i] == valid_detections_df$motusTagID[i - 1] &
         valid_detections_df$recvDeployName[i] != valid_detections_df$recvDeployName[i - 1] ) {
        calculate_distance(
          valid_detections_df$recvDeployLat[i], valid_detections_df$recvDeployLon[i],
          valid_detections_df$recvDeployLat[i - 1], valid_detections_df$recvDeployLon[i - 1]
        )
      } else {
        NA  # No distance calculation if receiver names are the same
      }
    }))
  )

valid_detections_df_SPPI <- valid_detections_df |>
  dplyr::select(ts, runID, motusTagID, tagDepLat, tagDepLon, group_id, recvDeployLat, recvDeployLon, recvDeployName, speciesEN, year, daysSinceDep, date, depDate100km, distBtwRecvKM, distance_from_previous_recv, time_diff_hours, timeDiff)

plotDataSPPI <- valid_detections_df_SPPI |>
  group_by(group_id) |>
  mutate(stopoverLengthDays = difftime(max(ts), min(ts), units = "days"),
         stopoverLengthRound = round(stopoverLengthDays, digits = 0),
         meanLat = mean(recvDeployLat),
         meanLon = mean(recvDeployLon),
         min_ts = as.Date(min(ts)), # adding the start and end points for plotting
         max_ts = as.Date(max(ts))) |>
  filter(recvDeployLat < 49) |>
  mutate(siteType = "Tag site")

```

```{r SPPI stopover maps, echo = FALSE, message = FALSE, warning = FALSE, fig.show = "hold", results = 'hide', fig.width=10, fig.height=8, fig.align='left', fig.cap="Possible stopover and overwintering locations for Sprague's Pipit. There were no detections spanning at least 24h within the species' known wintering range (out of range of this map).", fig.pos="H"}

xmin <- min(plotDataSPPI$meanLon, na.rm = TRUE) - 10
xmax <- max(plotDataSPPI$meanLon, na.rm = TRUE) + 10
ymin <- min(plotDataSPPI$meanLat, na.rm = TRUE) - 5
ymax <- max(plotDataSPPI$meanLat, na.rm = TRUE) + 5

ggplot(data = world) + 
  geom_sf(colour = "black") +
  #geom_sf(data = SPPI_winter_range, aes(fill = as.factor(season2)), color = NA, alpha = 0.3) + ignoring this layer since SPPI didn't have any stopovers >24h as far south as the wintering range
  geom_sf(data = lakes, colour = NA, fill = "white") +
  geom_sf(data = states_CanUS, colour = "black", fill = NA) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_bw() + 
  geom_point(data = plotDataSPPI, aes(x = tagDepLon, y = tagDepLat), colour = "black", fill = "red", shape = 23) +
  geom_point(data = filter(plotDataSPPI, stopoverLengthRound > 1), aes(x = meanLon, y = meanLat, size = as.numeric(stopoverLengthRound), fill = as.factor(motusTagID)), shape = 21) +
  labs(x = "", y = "", fill = "Tag ID") +
  geom_text(data = states_CanUS, aes(x = longitude, y = latitude, label = postal)) + 
  geom_label(data = plotDataSPPI, aes(x = tagDepLon, y = tagDepLat), label = "Tag Site", colour = "black", size = 3, hjust = 0.75, vjust = -0.5) +
  geom_label(data = plotDataSPPI, aes(x = meanLon, y = meanLat, label = paste("Date range:", min_ts, "-", max_ts)), color = "black", size = 3, vjust = -1) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  # Adjust the legend for stopoverLengthDays to range from 17 (shortest stopover) to 25 (longest stopover) by 2
  scale_size_continuous(name = "Stopover Length (days)", 
                        breaks = c(6),  # Define the breaks for the legend
                        labels = c(6),  # Define the labels for the legend
                        range = c(2, 10)) +
  scale_fill_manual(values = c("61024" = "#440154"))

```

\newpage
# References

Beason, R. C. (2020). Horned Lark (Eremophila alpestris), version 1.0. In Birds of the World (S. M. Billerman, Editor). Cornell Lab of Ornithology, Ithaca, NY, USA. https://doi-org.proxy.library.carleton.ca/10.2173/bow.horlar.01

Bleho, B., K. Ellison, D. P. Hill, and L. K. Gould (2020). Chestnut-collared Longspur (Calcarius ornatus), version 1.0. In Birds of the World (P. G. Rodewald, Editor). Cornell Lab of Ornithology, Ithaca, NY, USA. https://doi-org.proxy.library.carleton.ca/10.2173/bow.chclon.01

Davis, S. K., M. B. Robbins, and B. C. Dale (2020). Sprague's Pipit (Anthus spragueii), version 1.0. In Birds of the World (A. F. Poole, Editor). Cornell Lab of Ornithology, Ithaca, NY, USA. https://doi-org.proxy.library.carleton.ca/10.2173/bow.sprpip.01

Ellison, K., E. McKinnon, S. Zack, S. Olimb, R. Sparks, and E. Strasser (2017). Migration and winter distribution of the Chestnut-collared Longspur. Animal Migration (4): 37-40. https://doi.org/10.1515/ami-2017-0005 

Fink, D., T. Auer, A. Johnston, M. Strimas-Mackey, S. Ligocki, O. Robinson, W. Hochachka, L. Jaromczyk, C. Crowley, K. Dunham, A. Stillman, I. Davies, A. Rodewald, V. Ruiz-Gutierrez, C. Wood. 2023. eBird Status and Trends, Data Version: 2022; Released: 2023. Cornell Lab of Ornithology, Ithaca, New York. https://doi.org/10.2173/ebirdst.2022

